Clique and Constraint Models for Maximum Common (Connected) Subgraph Problems

### 2016-04-06 paper/paper.tex
More text and cites

Maximum common subgraph problems arise in biology and chemistry
\cite{DBLP:journals/jcamd/RaymondW02a,Ehrlich:2011}, computer vision ?? what should we cite?, ??
anything else?, both directly and as a way of measuring the similarity or difference between two
graphs \cite{DBLP:journals/prl/Bunke97,DBLP:journals/prl/FernandezV01,KriegeThesis}. We illustrate
this in \cref{figure:mcsexample}, and define the problem formally below.

\subsection{Definitions and Notation}

We will be working with graphs which may be directed, which may have labels associated with
vertices, edges, or both, and which may contain loops. To simplify notation, we will work with a
single definition throughout: we define a \emph{graph} $G = (V, \ell, e)$ to be a set of vertices
$V$, a vertex labelling function $\ell : V \rightarrow \mathbb{N}$, and an edge labelling function
$e : V \times V \rightarrow \mathbb{N} \cup \{ \bot \}$. To represent graphs without vertex (or
edge) labels, we just pick our favourite number and give it to every vertex (or edge). We also use a
distinct label $\bot$ to represent the \emph{lack of} an edge between two vertices; vertices $v$ and
$w$ are said to be \emph{adjacent} if $e(v, w) \ne \bot$ or $e(w, v) \ne \bot$. We write
$\operatorname{V}(G)$ for $V$, and $\operatorname{N}(G, v)$ for the \emph{undirected neighbourhood}
of a vertex $v$ (that is, the set of vertices adjacent to $v$); a vertex is \emph{isolated} if its
neighbourhood is empty. A \emph{path} between two vertices $v$ and $w$ is a sequence of $n$ vertices
$v = v_1, v_2, \ldots, v_n = w$ such that for each $m \in \{ 1 \ldots n - 1 \}$, $v_m$ is adjacent
to $v_{m + 1}$. A graph is \emph{connected} if a path exists between each distinct pair of vertices.

@@

There is also a non-induced version of these problems, which is commonly known as the \emph{maximum
common partial subgraph} problem: here, if $e(v, w) = \bot$, then $e_1(i_1(v), i_1(w))$ and
$e_2(i_2(v), i_2(w))$ may take any value, and we must maximise the number of non-$\bot$ vertex pairs
(i.e.\ edges) in $G$ instead (it is not entirely clear what the objective should be in directed
graphs, or when loops are present). We restrict ourselves to the induced version in this paper (??
we could do a journal version, because the clique side is rather involved\ldots).

@@

Another widely studied approach is to reduce the problem to finding a maximum clique (or sometimes,
equivalently, an independent set or a vertex cover ?? cite) in an association graph (which we describe
below). Conventional wisdom is that this approach is less effective in practice, but previous
experimental evaluations have used weak clique algorithms, or even maximal clique enumeration
algorithms. Maximum clique algorithms are an active research area
\cite{DBLP:conf/dmtcs/TomitaS03,DBLP:journals/jgo/TomitaK07,DBLP:conf/walcom/TomitaSHTW10,DBLP:journals/cor/SegundoRJ11,DBLP:journals/algorithms/Prosser12,DBLP:journals/ol/SegundoMRH13,DBLP:conf/ictai/LiFX13,DBLP:journals/cor/SegundoT14,DBLP:conf/lion/SegundoLB14,DBLP:conf/cp/McCreeshP14,DBLP:journals/jco/BatsynGMP14,DBLP:journals/cor/SegundoNB15,DBLP:conf/lion/NikolaevBS15,DBLP:conf/lion/LiJX15,DBLP:journals/jcc/KoncDTRJ12,DBLP:journals/algorithms/McCreeshP13,DBLP:journals/topc/McCreeshP15,DBLP:journals/cor/SegundoLP16},
and so the first half of this paper re-evaluates the association graph approach using a modern
algorithm.

(Other approaches have been tried, including dedicated algorithms ?? cite, mixed integer programming
\cite{DBLP:journals/anor/PivaS12}, and heuristics ?? cite; SAT encodings seem to struggle even for
subgraph isomorphism??  cite IJCAI.)

An interesting variant of the maximum common subgraph problem requires the selected subgraph to be
connected
\cite{DBLP:journals/tcs/Koch01,DBLP:journals/jcamd/RaymondW02a,DBLP:conf/mco/VismaraV08,Ehrlich:2011}.
Adding the connectedness requirement makes certain special cases solvable in polynomial time,
including outerplanar graphs of bounded degree \cite{DBLP:journals/algorithms/AkutsuT13} and trees
\cite{DBLP:journals/corr/DroschinskyKM16}, but the general case remains \NP-hard. The second half of
this paper looks at this variant on arbitrary graphs. In principle, for a constraint programming
approach, we may simply add a ``forms a connected graph'' global constraint to the model
\cite{Brown:2005,DBLP:conf/cp/DoomsDD05,DBLP:conf/cp/QuesadaRD05}. Alternatively, we could use a
special branching rule to maintain connectedness during search ?? cite. We see that using both
approaches together gives best results ?? does it?.  We also explain how to adapt this branching
rule to run inside a modern clique-like algorithm, to solve the problem via an association graph.

?? Summary, association graph is better when edges have labels.

@@

Explain association graph \cref{figure:association}, cite history, connections to microstructure
\cite{DBLP:conf/aaai/Jegou93a}.

>>>>

Maximum common subgraph problems arise in biology and chemistry
\cite{DBLP:journals/jcamd/RaymondW02a,Ehrlich:2011}, in computer vision
\cite{DBLP:journals/jair/CookH94}, in the analysis of source code
\cite{DBLP:journals/tkde/DjokoCH97}, binary programs \cite{DBLP:conf/icics/GaoRS08}, and circuit
designs \cite{DBLP:journals/jair/CookH94}, in character recognition problems \cite{SIWEILU1991617},
and in many other domains \cite{Shasha:2002:AAT:543613.543620}, both directly and as a way of
measuring the similarity or difference between two graphs
\cite{DBLP:journals/prl/Bunke97,DBLP:journals/prl/FernandezV01,KriegeThesis}. We illustrate two
variants of the problem in \cref{figure:mcsexample}.

Throughout this paper we will be working with graphs which may be directed, which may have labels
associated with vertices, edges, or both, and which may contain loops. To simplify notation, we will
work with a very broad definition: we define a \emph{graph} $G = (V, \ell, e)$ to be a set of
vertices $V$, a vertex labelling function $\ell : V \rightarrow \mathbb{N}$, and an edge labelling
function $e : V \times V \rightarrow \mathbb{N} \cup \{ \bot \}$. To represent graphs without vertex
(or edge) labels, we just pick our favourite number and give it to every vertex (or edge). We also
use a distinct label $\bot$ to represent the \emph{lack of} an edge between two vertices; vertices
$v$ and $w$ are said to be \emph{adjacent} if $e(v, w) \ne \bot$ or $e(w, v) \ne \bot$. We write
$\operatorname{V}(G)$ for $V$, and $\operatorname{N}(G, v)$ for the \emph{undirected neighbourhood}
of a vertex $v$ (that is, the set of vertices adjacent to $v$); a vertex is \emph{isolated} if its
neighbourhood is empty. A (weak) \emph{path} between two vertices $v$ and $w$ is a sequence of $n$
vertices $v = v_1, v_2, \ldots, v_n = w$ such that for each $m \in \{ 1 \ldots n - 1 \}$, $v_m$ is
adjacent to $v_{m + 1}$. A graph is (weakly) \emph{connected} if a path exists between each distinct
pair of vertices.

@@

There is also a non-induced version of these problems, which is commonly known as the \emph{maximum
common partial subgraph} problem: here, if $e(v, w) = \bot$, then $e_1(i_1(v), i_1(w))$ and
$e_2(i_2(v), i_2(w))$ may take any value, and we must maximise the number of non-$\bot$ vertex pairs
(i.e.\ edges) in $G$ instead (it is not entirely clear what the objective should be in directed
graphs, or when loops are present). We restrict ourselves to the induced version in this paper.

@@

Another widely studied approach ?? cite is to reduce the problem to finding a maximum clique (or sometimes,
equivalently, an independent set or a vertex cover ?? cite) in an association graph \cite{LeviG}; we describe
this below. Conventional wisdom is that this approach is less effective in practice, but previous
experimental evaluations have used weak clique algorithms, or even maximal clique enumeration
algorithms. Maximum clique algorithms are an active research area
\cite{DBLP:conf/dmtcs/TomitaS03,DBLP:journals/jgo/TomitaK07,DBLP:conf/walcom/TomitaSHTW10,DBLP:journals/cor/SegundoRJ11,DBLP:journals/algorithms/Prosser12,DBLP:journals/ol/SegundoMRH13,DBLP:conf/ictai/LiFX13,DBLP:journals/cor/SegundoT14,DBLP:conf/lion/SegundoLB14,DBLP:conf/cp/McCreeshP14,DBLP:journals/jco/BatsynGMP14,DBLP:journals/cor/SegundoNB15,DBLP:conf/lion/NikolaevBS15,DBLP:conf/lion/LiJX15,DBLP:journals/jcc/KoncDTRJ12,DBLP:journals/algorithms/McCreeshP13,DBLP:journals/topc/McCreeshP15,DBLP:journals/cor/SegundoLP16},
and so the first half of this paper re-evaluates the association graph approach using a modern
algorithm.

(Other approaches have been tried, including dedicated algorithms ?? cite, mixed integer programming
\cite{DBLP:journals/anor/PivaS12}, and heuristics \cite{DBLP:journals/jcisd/EnglertK15}; SAT
encodings seem to struggle even for subgraph isomorphism \cite{UpcomingIJCAIPaper}.)

An interesting variant of the maximum common subgraph problem requires the selected subgraph to be
connected
\cite{DBLP:journals/tcs/Koch01,DBLP:journals/jcamd/RaymondW02a,DBLP:conf/mco/VismaraV08,Ehrlich:2011}.
Adding the connectedness requirement makes certain special cases solvable in polynomial time,
including outerplanar graphs of bounded degree \cite{DBLP:journals/algorithms/AkutsuT13} and trees
\cite{DBLP:journals/corr/DroschinskyKM16}, but the general case remains \NP-hard. The second half of
this paper looks at this variant on arbitrary graphs. In principle, for a constraint programming
approach, we may simply add a ``forms a connected graph'' global constraint
\cite{Brown:2005,DBLP:conf/cp/DoomsDD05,DBLP:conf/cp/QuesadaRD05} to the model. Alternatively, we
could use a special branching rule to maintain connectedness during search ?? cite. Neither approach
seems directly viable with an association graph encoding---however, we show that it is possible to
adapt the combined branching and bounding rule used by modern clique algorithms to preserve
connectedness during search.

Our experiments suggest that each approach gives broadly similar performance, although the
association graph works better when graphs have edge labels (and we can explain why), and the
constraint programming approach is better when edges are unlabelled. For the constraint programming
model, using both a global constraint \emph{and} the branching rule in conjunction seems to be the
safest option---although it is not always fastest, it is very rarely substantially worse than just
using one or the other, and is often substantially better.

@@

Explain association graph \cref{figure:association}, cite history \cite{LeviG}, connections to
microstructure \cite{DBLP:conf/aaai/Jegou93a}.
